Staged Abstract Interpreters

### 2019-03-26 paper/10_intro.tex
section 3
section 4

\section{Introduction}

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections, which first identifies the binding-time of variables in
the program, i.e., they can be known whether statically or dynamically, and then
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-time is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP
for short) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the stage
annotations (i.e., binding-time annotations) to be explicit from the
programmers. The stage annotations can be either syntactic (as in
MetaML/MetaOCaml, through quasi-quote, escape, etc.) or type-based \todo{cite}.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we use the power function and type-based
annotations to introduce the idea of MSP:

@@

Now to concretely illustrate the idea of Futamura projection, consider that we
have an operational interpreter @eval@ of some language as following:

@@

where $e$ is the program of type @Expr@ and @arg@ is the input to program $e$,
and @eval@ produces a value.
Semantically, the interpreter satisfies $ \texttt{eval}(e)(arg) = [\![ e ]\!] arg$ for all
programs and inputs. Given a program $e_0$, by applying the first Futamura
projection, we may obtain a specialized interpreter
$\texttt{eval}_{\texttt{e0}}$ with respect to $e_0$ -- the so-called \textit{equivalent executable}. 
By definition of the interpreter, 
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $. If we generalize the
argument @arg@ to an environment (as well as a companion heap object, if for the
interest of using assignments) that maps free variables in $e$ to some values, 
the interpreter is a standard environment/store-passing interpreter,
where the specialization through MSP is also applicable.

Roughly at the same time when Futamura projections was visioned in 1970s,
\citeauthor{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fix-points \cite{DBLP:conf/popl/CousotC77}. However,
constructing sound abstract interpreters was considered abstruse and complicated
for a long time.
Recent progress such as Abstracting Abstract Machines (AAM)
uncovers a systematic method to derive sound abstract interpreters from their
concrete counterparts, where the soundness can be easily established by
examining the transformation of semantics.
The AAM approach also has been been applied to different variants of
definitional interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
  DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.

@@

Now we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by
approximation, we can refactor interpreters to static analyzers. As an
intellectual quest, we should naturally ask -- starting from an interpreter,
\textit{can we derive a sound abstract interpreter while it produces fast code
that does the analysis?}
Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to a program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the compiled one leaves the
abstract environment $\widehat{\rho}$ as remaining input.

\begin{figure}[h]
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=3em,minimum width=2em]
  {
    \begin{smallmatrix} \text{unstaged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{abstract} \\ \text{interpreters} \end{smallmatrix} \\
      \begin{smallmatrix} \text{unstaged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} & \begin{smallmatrix} \text{staged} \\ \text{concrete} \\ \text{interpreters} \end{smallmatrix} \\ };
  \path[-stealth]
    (m-1-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-1-2)
    (m-2-1) edge node [left, font=\scriptsize] {$\alpha$} (m-1-1)
    (m-2-1) edge node [below, font=\scriptsize] {$\updownarrows$} (m-2-2)
    (m-2-2) edge node [right, font=\scriptsize] {$\alpha$} (m-1-2);
  \end{tikzpicture}
  \caption{The confluence of specialization and approximation.}
  \label{confluence}
\end{figure}

@@

We show that these well-known abstractions can be combined smoothly, and more
importantly the price is almost free. Through staging, the major overhead of
abstractions -- from the monadic layers -- is completely eliminated in the
generated code. However, we do not lose any advantages of this high-level
programming style (e.g., the extensibility); the meaning of the program also
remains the same, particularly, the abstract semantics that does the analysis
when instantiated as an abstract interpreter.

@@

Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter one while the former one is untouched.
In the literature, specializing analysis with respect to a program by partial
evaluation \cite{damian1999partial, amtoft1999partial,
Boucher:1996:ACN:647473.727587, ashley:practical} is not a new, despite the fact
that most of the previous work are ad-hoc on a particular analysis or requiring
significant changes to the analyzer. For example, abstract compilation
\cite{Boucher:1996:ACN:647473.727587} requires the whole analyzer to be
rewritten as closure generation from. In this paper, we advocate a systematic
approach to meta-programming that achieves the performance goal with less
intrusive changes; in fact, a case study (Section~\ref{cs_ac}) comparing with
abstract compilation shows, by utilizing type-based stage annotations, the
analyzer program does not need any changes.

\paragraph{Modularity via Meta-Programming}

@@

>>>>

\section{Introduction} \label{intro}

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections, which first identifies the binding-time of variables in
the program, i.e., they can be known whether statically or dynamically, then
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-times is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP
for short) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the stage
annotations (i.e., binding-time annotations) to be explicit from the
programmers. The stage annotations can be either syntactic (as in
MetaML/MetaOCaml, through quasi-quote, escape, etc.) or type-based \todo{cite}.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we use the power function and type-based
annotations to introduce the idea of MSP:

@@

Now to concretely illustrate the idea of Futamura projection, consider that we
have an operational interpreter @eval@ of some language:

@@

where $e$ is the program of type @Expr@ and @arg@ is the input to program $e$,
and @eval@ produces a value.
Semantically, the interpreter satisfies $ \texttt{eval}(e)(arg) = [\![ e ]\!] arg$ for all
programs and inputs. Given a program $e_0$, by applying the first Futamura
projection, we may obtain a specialized interpreter
$\texttt{eval}_{\texttt{e0}}$ with respect to $e_0$ -- the so-called \textit{equivalent executable}. 
By definition of the interpreter, 
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $. If we generalize the
argument @arg@ to an environment (as well as an accompanying heap object, if for
the interest of using assignments) that maps free variables in $e$ to some
values, the interpreter is a standard environment-passing interpreter,
where the specialization through MSP is also applicable.

Roughly at the same time when Futamura projections was visioned in 1970s,
\citet{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fix-points. However, constructing sound abstract interpreters
was considered abstruse and complicated for a long time.
Recent progress such as Abstracting Abstract Machines (AAM)
\cite{DBLP:conf/icfp/HornM10} uncovers a systematic method to derive sound
abstract interpreters from their concrete counterparts, where the soundness can
be easily established by examining the transformation of semantics. The AAM
approach also has been been applied to different variants of definitional
interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.

@@

Now we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by
approximation, we can refactor interpreters to static analyzers. As an
intellectual quest, we should naturally ask -- starting from an abstract
interpreter, \textit{can we derive a sound abstract interpreter while it
produces fast code that does the analysis?}
Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to some program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the compiled one leaves the
abstract environment $\widehat{\rho}$ as remaining input.

\vspace{-1em}
\begin{figure}[h]
  \begin{tikzpicture}
  \matrix (m) [matrix of math nodes,row sep=3em,column sep=3em,minimum width=2em]
  {
    \begin{smallmatrix} \text{unstaged} \\ \text{abstract} \\ \text{interpreters} \\ (\text{Sec.}~\ref{unstaged_abs}) \end{smallmatrix} &
    \begin{smallmatrix} \text{staged} \\ \text{abstract} \\ \text{interpreters} \\ (\text{Sec.}~\ref{sai}) \end{smallmatrix} \\
    \begin{smallmatrix} \text{unstaged} \\ \text{concrete} \\ \text{interpreters} \\ (\text{Sec.}~\ref{prelim}) \end{smallmatrix} &
    \begin{smallmatrix} \text{staged} \\ \text{concrete} \\ \text{interpreters} \\ (\text{Sec.}~\ref{stagedinterp}) \end{smallmatrix} \\ };
  \path[-stealth]
    (m-1-1) edge node [above, font=\scriptsize] {$\updownarrows$} (m-1-2)
    (m-2-1) edge node [left, font=\scriptsize] {$\alpha$} (m-1-1)
    (m-2-1) edge node [below, font=\scriptsize] {$\updownarrows$} (m-2-2)
    (m-2-2) edge node [right, font=\scriptsize] {$\alpha$} (m-1-2);
  \end{tikzpicture}
  \caption{The confluence of specialization and approximation.}
  \label{confluence}
\end{figure}
\vspace{-1em}

@@

We show that these well-known abstractions can be combined smoothly, and more
importantly the price is almost free. Through staging, the major overhead of
abstractions -- from the monadic layers -- is completely eliminated in the
generated code. \todo{put some numbers} However, we do not lose any advantages of this high-level
programming style (e.g., modularity and extensibility); the meaning of the program also
remains the same, particularly, the abstract semantics that does the analysis
when instantiated as an abstract interpreter.

@@

Besides the intellectual merit shown in the confluence diagram, this paper
contributes to a practical problem in static analysis: \textit{can we speed up a
static analyzer without any intrusive changes and soundness compromises?} Static
analysis is known as a trade-off game between \textit{precision} and
\textit{efficiency}, we argue that applying meta-programming (multi-stage
programming, in particular) with abstract interpreters is an effective approach
to improve the latter one while the former one is untouched.
In the literature, specializing analysis with respect to the input program
\cite{damian1999partial, amtoft1999partial, Boucher:1996:ACN:647473.727587,
ashley:practical} is not a new, despite the fact that most of the previous work
are ad-hoc on a particular analysis or requiring significant changes to the
analyzer. For example, abstract compilation is one of the techniques
\cite{Boucher:1996:ACN:647473.727587}. But it requires the whole analyzer to be
rewritten to the closure generation from. In this paper, we advocate a systematic
approach to meta-programming that achieves the performance goal with less
intrusive changes; in fact, a case study (Section~\ref{cs_ac}) comparing our
approach with abstract compilation shows that, by utilizing type-based stage
annotations, the analyzer program does not need any changes.

\paragraph{Modularity via Meta-Programming} \todo{compress this part}

@@

\todo{mention LMS somewhere}

### 2019-03-26 paper/30_stageconc.tex
section 4

Readers may notice that the conversion between unstaged monads and monads is 
simply changing the unstaged data to staged data, which in fact can be achieved without
changing the implementation of monads. This is true so far, but as we will see
when introducing the nondeterminism monad (@ListT@) for abstract interpreters, as
the whole list will be a next-stage value instead of merely the elements in the list.
For this reason, we explicitly distinguish and present the two version of
monads, as a more consistent approach.

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to staged version -- we just need to let them works on staged data.
As we mentioned before, we elaborate how to handle functions and applications in
detail. One of the desired goal is to eliminate the monadic layers through
staging, this becomes tricky for clsoures, because we need to actually compile
the lambda term instead of generate a @CloV@ value. To make this happen, when
closing the lambda term with the environment, we need to \textit{collpase} the
monads to normal values. The following function @close@ illustrates the idea:
@f@ is a current stage function but takes two next-stage values @v@ and
$\sigma$, inside of which, we eagerly collapse the monads to values by
@ev(e)(ρ_*)(σ_*)@, which has type @Rep[(Value, Store)]@.
Then we generate an intermediate representation of the compiled closure using
@emit_compiled_clo@. The intermediate representation stands for a next-stage
value, and contains the compiled function, which will be generated as a
next-stage function.

@@

The method @emit_compiled_clo@ and @emit_ap_clo@ mentioned above produces IR
node in LMS. We define these IR node as follow: \note{Def, quote are omitted}

>>>>

Readers may notice that the conversion between unstaged monads and monads is 
simply changing the unstaged data to staged data, which in fact can be achieved without
changing the implementation of monads. This is true so far, but as we will see
it is not that easy when introducing the nondeterminism monad (@ListT@) for
abstract interpreters, as the whole list will be a next-stage value instead of
merely the elements in the list.
For this reason, we explicitly distinguish and present the two version of
monads, as a more consistent approach.

\paragraph{Primitive Operations} Most of the primitive operations can be easily
translated to staged version -- we just need to let them works on staged data.
As we mentioned before, we elaborate how to handle functions and applications in
detail. One of the desired goal is to eliminate the monadic layers through
staging, this becomes tricky for clsoures, because we need to actually compile
the lambda term instead of generating a @CloV@ value. To make this happen, when
closing the lambda term with the environment, we need to staging into the body
expression by \textit{collpasing} the @AnsM@ monads to normal values.
The following function @close@ illustrates the idea:
@f@ is a current stage function but takes two next-stage values @v@ and
$\sigma$, inside of which, we eagerly collapse the monads to values by
@ev(e)(ρ_*)(σ_*)@, which has type @Rep[(Value, Store)]@.
Then we generate an intermediate representation of the compiled closure using
@emit_compiled_clo@. The intermediate representation stands for a next-stage
value, and contains the compiled function, which will be generated as a
next-stage function.

@@

The method @emit_compiled_clo@ and @emit_ap_clo@ mentioned above produces IR
node in LMS. We define these IR node as follow: \note{Exp, Def, quote are omitted}

### 2019-03-26 paper/40_absinterp.tex
section 4

After seeing both unstaged and staged concrete interpreters, now we trun our
focus on abstract interpreters under the same framework. We will first present a
lattice framework with binding-time annotations, but for the purpose of clearly
illustrating the idea, we just use simple abstract domains such as power set
lattice or point-wise map lattice. Then we show the abstract components such as
abstract environments and abstract stores.
The abstract interpreter we present in this section is similar to
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s, it is also
context/path/flow-insensitive by our choice.  In Section \ref{cfa}, we will
instantiate a context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitive by varying the monads, which is also applicable to our unstaged or
staged abstract interpreter. \todo{numerical/imp?}

@@

\begin{lstlisting}
  trait SPLattice[S, R[_]] {
    val $\top$: R[S];  val $\bot$: R[S]
    def $\sqsubseteq$(l1: R[S], l2: R[S]): R[Boolean]
    def $\sqcup$(l1: R[S], l2: R[S]): R[S]
    def $\sqcap$(l1: R[S], l2: R[S]): R[S]
  }
  trait Lattice[S] extends SPLattice[S, NoRep]
\end{lstlisting}

@@

\begin{lstlisting}
  def SetLattice[T]: Lattice[Set[T]] = new Lattice[Set[T]] {
    lazy val $\bot$: Set[T] = Set[T]()
    lazy val $\top$: Set[T] = throw new NotImplementedError("No representation for top")
    def $\sqsubseteq$(l1: Set[T], l2: Set[T]): Boolean = l1 subsetOf l2
    def $\sqcup$(l1: Set[T], l2: Set[T]): Set[T] = l1 union l2
    def $\sqcap$(l1: Set[T], l2: Set[T]): Set[T] = l1 intersect l2
  }
\end{lstlisting}

@@

A cache is a mapping from configurations to sets of value-store pairs, where the
configuration is a triple of current expression being evaluated, environment and store.

\begin{lstlisting}
  type Config = (Expr, Env, Store); type Cache = Map[Config, Set[(Value, Store)]]
  type Result = (R[List[(Value, Store)]], R[Cache])
\end{lstlisting}

\paragraph{Monads for Abstract Interpretation} nondetT

\paragraph{Primitive Operations}

@@

\paragraph{Caching and Fixpoint Iteration}

\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]
  def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def update_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

>>>>

After seeing both unstaged and staged concrete interpreters, now we trun our
focus on abstract interpreters under the same framework. We will first present a
lattice framework with binding-time annotations, but for the purpose of clearly
illustrating the idea, we just use simple abstract domains such as power set
lattice or point-wise map lattice. Then we show the abstract components such as
abstract environments and abstract stores.
The abstract interpreter we present in this section is similar to
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s, it is also
context/path/flow-insensitive by our choice.  In Section \ref{cfa}, we will
instantiate a context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our unstaged
or staged abstract interpreter.

@@

\begin{lstlisting}
  trait SPLattice[S, R[_]] {
    val ⊤: R[S];  val ⊥: R[S]
    def ⊑(l1: R[S], l2: R[S]): R[Boolean]
    def ⊔(l1: R[S], l2: R[S]): R[S]
    def ⊓(l1: R[S], l2: R[S]): R[S]
  }
  trait Lattice[S] extends SPLattice[S, NoRep]
\end{lstlisting}

@@

\begin{lstlisting}
  def SetLattice[T]: Lattice[Set[T]] = new Lattice[Set[T]] {
    lazy val ⊤: Set[T] = Set[T]()
    lazy val ⊥: Set[T] = throw new NotImplementedError("No representation for top")
    def ⊑(l1: Set[T], l2: Set[T]): Boolean = l1 subsetOf  l2
    def ⊔(l1: Set[T], l2: Set[T]): Set[T]  = l1 union     l2
    def ⊓(l1: Set[T], l2: Set[T]): Set[T]  = l1 intersect l2
  }
\end{lstlisting}

@@

To prevent the abstract interpreter from falling into nontermination, the ADI
approach uses a two-cache mechanism. Here we first provides the type
definitions, and later we will describe the algorithm in detail. A @Cache@ is a
mapping from configurations @Config@ to sets of value-store pairs, where the
configuration is a triple of current expression being evaluated, environment and
store.

\begin{lstlisting}
  type Config = (Expr, Env, Store); type Cache = Map[Config, Set[(Value, Store)]]
\end{lstlisting}

As of now, the store is a mapping from addresses to sets of abstract values.
Therefore, when applying a function, for instance @f(a)@, we may retrieve a set
of closures by dereferencing the address of @f@. We need to apply all of the
target closures, where the nondeterminism is introduced. Another nondeterminism
happens when evaluating conditionals, we have to evaluate both branches if the
condition can not be determined. In other words, we will have multiple computing
results.

\begin{lstlisting}
  type Result = (R[List[(Value, Store)]], R[Cache])
\end{lstlisting}

This oberservation leads us to the type definition of @Result@: in concrete setting,
the @Result@ is a pair of @Value@ and @Store@, now we use a list of such pairs
@List[(Value, Store)]@ to represent multiple results. The results is also linked
with the resulting cache used in this iteration.
Note that the abstract binding-time annotation is wrapped to the two objects in
the @Result@ pair, but not the whole pair. Because we statically know the result
is a tuple of two elements, but we don't know how many value-store pairs we
would have, neither the content of the cache.

\paragraph{Monads for Abstract Interpretation} Now we incorporate the
nondeterminism and caching into the monad stack. For concrete evaluation, we
already have a reader effect for environments and a state effect for stores. We
add three more effects based on it: the nondeterminism effect is represented by
the @ListT[M[_], A]@ monad transformer, inside of which, we have a @ReaderT@ for
a @in@ cache, and a @StateT@ for an @out@ cache.
\todo{highlight where changed}

\begin{lstlisting}
  trait AbstractSemantics extends AbstractComponents {
    type R[T] = T
    type AnsM[T] = ReaderT[StateT[ListT[ReaderT[StateT[IdM, Cache, ?], Cache, ?], ?], Store, ?], Env, T]
    ...
  }
\end{lstlisting}

The sketch implementation of @ListT@ is shown below. The value @run@ it encapsulates
is of type @M[List[A]]@, where @M[_]@ is the inner monad, and @A@ the element
type of the list.

\begin{lstlisting}
  case class ListT[M[_]: Monad, A](run: M[List[A]]) { ... }
\end{lstlisting}

\paragraph{Primitive Operations} \todo{wip}

@@

\paragraph{Caching and Fixpoint Iteration} Caching

\begin{lstlisting}
  def ask_in_cache: AnsM[Cache]; def get_out_cache: AnsM[Cache]
  def put_out_cache(out: R[Cache]): AnsM[Unit]
  def set_out_cache(cfg: R[Config], vs: R[(Value, Store)]): AnsM[Unit]
\end{lstlisting}

### 2019-03-26 paper/50_stageabs.tex
section 4

\subsection{Modular Analysis for Free}

One of the challenges of modern static analysis is program usually depends on
large libraries programs~\cite{toman_et_al:LIPIcs:2017:7121}. Can we analyze
programs and libraries separately and reuse the result without losing precision?
Then we can reduce part of the overhead of repeatedly analyzing libraries for
different programs. Indeed, some static analyzers compute summary for a function
or a module, which can be reused later, however they are mostly too conservative
or unsound, which both lead to imprecision.

The specialization of abstract interpreter provides a chance to obtain such
partial analysis result in a mechanized way, but still keeps the analysis sound.
As we see when compiling the closures, we can specialize the abstract
interpreter with respect the body expression of the lambda term without knowing
the actual argument. The programs with some unknown variables are open programs,
which is exactly the case if we want to analyze programs in a modular way.

For a concrete analysis, for example, $k$-CFA ($k$ > 0) is naturally a
whole-program analysis, because it is inter-procedural and needs the last $k$
calling contexts to distinguish different call sites, where the calling contexts
are dynamic values. However, it is possible to analyze programs (libraries)
separately through specializing an abstract interpreter that generates the
analysis as the next-stage program and leave the unavailable programs and
calling contexts as dynamic parameters, and then install these contexts when we
have the whole program.

\iffalse
Another perspective: programs are data for an abstract interpreter, so if we
have $n$ programs, then maybe there can be $n$ stages. Probably we can analyze
first $m$ programs, and generate a residual abstract interpreter waiting for the
rest $(n-m)$ programs. These $(n-m)$ programs might be (abstract) arguments for
the first $n$ programs, and the abstract interpreter itself might be a partial
abstract interpreter.
\fi

@@

Our staging schema works well in theory, but would suffer from code explosion,
and possibly runtime GC overhead when analyzing (specializing) large programs.
We present several optimizations that largely mitigate these issues. Still,
these optimizations do not need to change the generic interpreter.

@@

\paragraph{Are Monads Necessary?} No. One can always inline the monads and
obtain an abstract interpreter in continuation-passing style, or simply use
explicit side-effects such mutation to implement the artifact with the same
abstract semantics. In either cases, we can still apply the staging schema to
abstract interpreter. As an evidence, we provide an staged abstract interpreter
written in direct-style in the accompanying artifact.

@@

What we implemented is a big-step, compositional abstract interpreter in monadic
style, where "compositional" means that every recursive call of our abstract
interpreter is applied to proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as the termination of
specialization procedure. It is also possible to specialize small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} --
\citet{Johnson:2013:OAA:2500365.2500604} implemented it for
optimizating Abstract Abstract Machines. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering efforts.

@@

As one of the merit of our approach, the staging does not compromise any
soundness, because we only change the underlying monads and data structures to
the staged version, not any part of the abstract semantics. Based on the
assumption that LMS preserves the equivalence during staging, we are confident
that the staged abstract interpreter does the same analysis as the unstaged one.

>>>>

@@

Our staging schema works well in theory, but would suffer from code explosion,
and possibly runtime GC overhead when analyzing (specializing) large programs.
We present several optimizations that largely mitigate these issues. Still,
implementing these optimizations do not need to change the generic interpreter.

@@

\paragraph{Are Monads Necessary?} No. One can always inline the monads and
obtain an abstract interpreter in continuation-passing style, or simply use
explicit side-effects such mutation to implement the artifact with the same
abstract semantics. In either cases, we can still apply the staging schema to
the abstract interpreter. As an evidence, we provide an staged abstract
interpreter written in direct-style in the accompanying artifact.

@@

What we implemented is a big-step, compositional abstract interpreter in monadic
style, where \textit{compositional} means that every recursive call of our abstract
interpreter is applied to proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as the termination of
specialization procedure. It is also possible to specialize small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} --
\citet{Johnson:2013:OAA:2500365.2500604} implemented it for
optimizating Abstract Abstract Machines. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering efforts.

@@

As one of the merit of our approach, the staging does not compromise any
soundness, because we only change the underlying monads and data structures to
the staged version, not any part of the abstract semantics. Based on the
assumption that MSP system and staged data structure preserve the equivalence
during staging, we are confident that the staged abstract interpreter does the
same analysis as the unstaged one.

### 2019-03-26 paper/60_cases.tex
section 4

\section{Case Study} \label{cases_study}

@@

%\subsubsection{Mixed Sensitivity}

%using different $k$ for $k$-CFA

>>>>

\section{Case Studies} \label{cases_study}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Modular Analysis for Free}

% \subsubsection{Mixed Sensitivity}
% using different k?

One of the challenges of modern static analysis is program usually depends on
large libraries programs~\cite{toman_et_al:LIPIcs:2017:7121}. Can we analyze
programs and libraries separately and reuse the result without losing precision?
Then we can reduce part of the overhead of repeatedly analyzing libraries for
different programs. Indeed, some static analyzers compute summary for a function
or a module, which can be reused later, however they are mostly too conservative
or unsound, which both lead to imprecision.

The specialization of abstract interpreter provides a chance to obtain such
partial analysis result in a mechanized way, but still keeps the analysis sound.
As we see when compiling the closures, we can specialize the abstract
interpreter with respect the body expression of the lambda term without knowing
the actual argument. The programs with some unknown variables are open programs,
which is exactly the case if we want to analyze programs in a modular way.

For a concrete analysis, for example, $k$-CFA ($k$ > 0) is naturally a
whole-program analysis, because it is inter-procedural and needs the last $k$
calling contexts to distinguish different call sites, where the calling contexts
are dynamic values. However, it is possible to analyze programs (libraries)
separately through specializing an abstract interpreter that generates the
analysis as the next-stage program and leave the unavailable programs and
calling contexts as dynamic parameters, and then install these contexts when we
have the whole program.

\iffalse
Another perspective: programs are data for an abstract interpreter, so if we
have $n$ programs, then maybe there can be $n$ stages. Probably we can analyze
first $m$ programs, and generate a residual abstract interpreter waiting for the
rest $(n-m)$ programs. These $(n-m)$ programs might be (abstract) arguments for
the first $n$ programs, and the abstract interpreter itself might be a partial
abstract interpreter.
\fi
